---
title: "Class Examples"
author: "Alex Vand"
date: "4/7/2022"
output: html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

# 2022-04-07

```{r}
percent_change_gnp = function(percent_change_unemployment_rate){
  percent_change_gnp <- 0.856 - 1.827 * percent_change_unemployment_rate
  return(percent_change_gnp)
}
```



```{r}
percent_change_gnp(3)
```



```{r}
percent_change_gnp = function(percent_change_unemployment_rate){
  percent_change_gnp <- 0.856 - 10 * percent_change_unemployment_rate
  return(percent_change_gnp)
}
```


```{r}
percent_change_gnp(3)
```




```{r}
percent_change_gnp = function(percent_change_unemployment_rate,
                              intercept,
                              slope){
  percent_change_gnp <- intercept + slope * percent_change_unemployment_rate
  return(percent_change_gnp)
}
```



```{r}
percent_change_gnp(3, 0.856, -1.827)
```




# 2022-04-12

```{r}
source("R/compute_diversity.R")
```


```{r}
flower_species <- c("iris", "daisy", "rose", "tulip", "lily", "sunflower")
```


# 2022-04-19

```{r}
library(pse)
```


# 2022-04-26

```{r}
library(sensitivity)
```




#2022-04-28

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(purrr)
```

Implementing Dynamic Models in R

Dynamic models always involves derivatives (equations that express how things change from time step to time step or place to place )


Implement population **growth** as a derivative - a model of population change

```{r, echo=TRUE}
# derivatives
# note that we include time here but we don't use it; we will need this later
source("R/dexppop.R")

# see how it works
dexppop(P=20, r=0.01)

#what is this?

# notices this is the same as
dexppop(t=100,P=20, r=0.01)

# lets look at this for a range of initial populations
pops = seq(from=1, to=100)
tmp = pops %>% map(~dexppop( time=0,r=0.01, P=.x))
pchange = unlist(tmp)
pdyn = data.frame(pops, pchange)

ggplot(pdyn, aes(pops, pchange))+geom_point(col="green", size=1.5)
# why is this a straight line?
# how many new individuals are born at each population level
# try this - add a carrying capacity (dP/dt = 0 if P > carryingcapacity)
```

Integration

What if we wanted to look at population in 20 years given an initial condition

Two options

* explicit solution to differential equation is known; e.g. you can integrate both sides of the equation! Not always possible but lets look at a case where it is possible

* must be solved by iteration; this is what we do when we can't integrate both sides 

Explicit Solution is available


```{r}
# explicit
source("R/exppop.R")
# gives population after any time given an initial population
# 20 rabbits, growth rate of 0.01 how many in 30 years
exppop(T=30, P0=20, r=0.01, K=1000)

# if we want to see how population evolves over time - generate a time series by running our model for each point in time
initialrabbits = 20
years = seq(from=1, to=100, by=2)
Ptime = years %>% map_dbl(~exppop( P0=initialrabbits, r=0.01, K=1000, T=.x))

# keep track of what times we ran
Ptime = data.frame(P=Ptime, years=years)
ggplot(Ptime, aes(years,P))+geom_point()+labs(x="years",y="Rabbit Population")

# try generating results for maximum and minimum possible r values to compare (guess at what you think)
max_r = 0.1
min_r = 0.01
K = 1000
tmp = years %>% map_dbl(~exppop(r=max_r, P0=initialrabbits, K=K, T=.x))
Ptime$Pmaxr = tmp
tmp = years %>% map_dbl(~exppop(r=min_r, P0=initialrabbits, K=K, T=.x))
Ptime$Pminr = tmp
head(Ptime)

Ptimep = Ptime %>% gather(key="r",value="P",-years)
ggplot(Ptimep, aes(years,P, col=r))+geom_point()+labs(x="years",y="Rabbit Population")
# notice how populatin becomes unstable for high growth rates! 
```

# 2022-05-03

```{r}
source("R/dexppop.R")

dexppop
```


```{r}
library(tidyverse)
library(deSolve)
initialrabbits = 20
years = seq(from=1, to=100, by=2)

# run the solver
Ptime = ode(y=initialrabbits, times=years, func=dexppop,parms=c(0.01))
head(Ptime)
```

additional parameters 

You can play a bit with changing your function to something that you can't integrate "by hand"

BUT we might want more parameters

to work with ODE,
parameters must all be input as a single list; similar to how we return multiple outputs from a function

see example below..lets add a carrying capacity

R code with carrying capacity

```{r}
source("R/dexppop_play.R")
dexppop
# create parameter list
initalrabbits=2
newparms = list(r=0.03, carry_capacity=300)
#apply solver
results=ode(initialrabbits, years, dexppop_play,newparms)
head(results)
# add more meaningful names
colnames(results)=c("year","P")
# plot
ggplot(as.data.frame(results), aes(year,P))+geom_point()+labs(y="Population", "years")
# try again with different parameters
alternativeparms = list(r=0.04, carry_capacity=500)
results2=ode(initialrabbits, years, dexppop_play,alternativeparms)
# look at results
head(results2)
colnames(results2)=c("year","P_parm2")
# plot
ggplot(as.data.frame(results2), aes(year,P_parm2))+geom_point()+labs(y="Population", "years")
# compare by combining into a single data frame
both = inner_join(as.data.frame(results), as.data.frame(results2))
both_p = both %>% gather(key="model", value="P", -year)
ggplot(both_p, aes(year,P, col=model))+geom_point()+labs(y="Population", "years")
# try playing on your own - modify the function in some way
```

Difference Equations 

What is ODE doing? (iterating in 'smart ways')

Similar to "difference equations"

Difference Equations

Population models can be discrete (rather than continuous)
So we could implement them as difference equations and iterate


```{r}
#iterate
source("R/discrete_logistic_pop.R")
# notice how a for loop is used to iterate
# how many rabbits after 50 years given a growth of 0.1
# starting with 1 rabbit - but a carrying capcity of 500
discrete_logistic_pop
discrete_logistic_pop(P0=1, r=0.05, K=200, T=50)
# save results
discrete_result = discrete_logistic_pop(P0=1, r=0.05, K=200, T=50)
# lets also keep the parameters for use later
P0=1
r=0.05
K=200
T=50
```

Differential Equation, Difference (Iteration by hand) comparison

Remember we have 3 ways now to calculate population

analytical solution - based on integration (exppop.R) BEST

using an ode solver for numerical approximation (exppop_play.R)

numerical integration using in discrete steps (discrete_logistic_pop.R)

```{r}
#closed
source("R/exppop.R")
exppop(P0=P0, r=r, K=K, T=T)
analytic_result = exppop(P0=P0, r=r, K=K, T=T)
analytic_result
discrete_result
# why are they different
# look at trajectories
growth_result = data.frame(time=seq(from=1,to=100))
growth_result$Panalytic = growth_result$time %>% map_dbl(~exppop( P0=1,r=0.05, K=200,T=.x ))
growth_result$Pdiscrete = growth_result$time %>% map_dbl(~discrete_logistic_pop( P0=1,r=0.05, K=200,T=.x ))
tmp = growth_result %>% gather(key="Ptype",value="P",-time)
ggplot(tmp, aes(time,P, col=Ptype))+geom_point()
# try running them for longer time periods to see what happens 
# change the value of r, K , P0 - see how it effects the results
```

Compare analytical, difference and ODE

Finally look at continuous derivative using ODE solve
Needs
*initial conditions
*differential equation
*parameters

```{r}
source("R/dexppop_play.R")
dexppop_play
# set up using the same parameters
pcompare = list(r=r,carry_capacity=K)
# now run our ODE solver
result = ode(P0, growth_result$time, dexppop_play, pcompare)
head(result)
# we already have time - so just extract population
growth_result$Pdifferential = result[,2]
# compare all 3 approaches
tmp = growth_result %>% gather(key="Ptype",value="P",-time)
ggplot(tmp, aes(time,P, col=Ptype))+geom_point()
# notice Pdifferential is closer to Panalytic than Pdiscrete
```



Diffusion


```{r}
source("R/diffusion.R")
# run our diffusion model (iterative difference equation) with initial concentration of 10, for 8 timestep (size 1m), and 10 space steps (size 1s)
# using diffusion parameters 0.5 s/m2, 10 m2
result = diff1(initialC=10, nx=10, dx=1, nt=8, dt=1, D=0.5, area=10)
# a list is returned with our 3 data frames for concentration (conc), qin and qout
result
# used filled contour to plot results
head(result$conc)
filled.contour(result$conc, xlab="Time", ylab="Distance")
# or if you prefer this orientation (Distance on x axis)
filled.contour(t(result$conc), ylab="Time", xlab="Distance")
```

Change parameters (diffusivity D, and space and time steps (dx, dt))

```{r}
# changes diffusivity and other parameters particularly
# diffusivity, dx and dt
#res=diff1(initialC=10,nx=10,dx=1,nt=10,dt=30,D=0.006,area=1)

#compare above res and below
res=diff1(initialC=10, nx=10, dx=1, nt=8, dt=1, D=0.5, area=10)

filled.contour(res$conc, xlab="Time", ylab="Distance")
# we can also see how much material moved from place to place each time step
filled.contour(res$qin, xlab="Time", ylab="Distance")
# play with time step, space step and parameters
```




```{r}
# changes diffusivity and other parameters particularly
# diffusivity, dx and dt
res=diff1(initialC=10,nx=10,dx=1,nt=10,dt=30,D=0.006,area=1)
#long dragged out is original

#compare above res and below
#res=diff1(initialC=10, nx=10, dx=1, nt=10, dt=30, D=006, area=1)

filled.contour(res$conc, xlab="Time", ylab="Distance")
# we can also see how much material moved from place to place each time step
filled.contour(res$qin, xlab="Time", ylab="Distance")
# play with time step, space step and parameters
```




# 2022-05-12

```{r}
source("R/av_dharvest.R")
```

```{r}
# given some forest characteristics - lets look at derivatives under different harvest rates
lowHrate = 10
gps = list(harv=lowHrate, K=1000, r=0.05)

# look at the derivative over a range of forest sizes

findstable = data.frame(Ccurr=seq(from=1, to=1000, by=10))
# notice use of Time=NULL, and map to compute derivative
# for different values of forest biomass
findstable$dervHlow= unlist(findstable$Ccurr %>% map(~dharvest(parms=gps, Time=NULL, P=.x) ))
                                                  
ggplot(findstable, aes(Ccurr, dervHlow))+geom_point()+geom_hline(yintercept = 0, col="red")+
  labs(y="Derivative\n (Rate of change of forest carbon) (kg/C/year)", x="Current Forest Carbon Stock (kgC)")
```

